---
output: github_document
---

<!-- 
README.md is generated from README.Rmd. Please edit that file
rmarkdown::render(
  "README.Rmd", output_format = rmarkdown::github_document(html_preview = FALSE)) 
-->


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/dplyrUtil)](http://cran.r-project.org/package=dplyrUtil)
[![Travis-CI Build Status](https://travis-ci.org/bgctw/dplyrUtil.svg?branch=master)](https://travis-ci.org/bgctw/dplyrUtil)


## Overview

The `dplyrUtil` package provides utilities that deal with common task and 
with unequal factor levels when using dplyr.

Common tasks when using dplyr are 

- split-map-combine
- mutate acting only on the rows satisfying the condition
- left_join with before dropping columns that would be duplicated

## Installation

```{r, eval = FALSE}
# From CRAN (in future)
#install.packages("dplyrUtil")

# Or the the development version from GitHub:
# install.packages("devtools")
devtools::install_github("bgctw/dplyrUtil")
```

## Usage

```{r}
require(dplyr, quietly = TRUE)
require(tidyr, quietly = TRUE)
require(dplyrUtil, quietly = TRUE)
```


Example of convenient split-map-combine with example dataset:
```{r}
ds <- tibble(
  food = c("banana","strawberry","pear","bread","corn flakes")
  , kind = factor(c(rep("fruit",3), rep("cereal",2)))
  )
ds
```

```{r}
fSub <- function(dss,kind){ 
  dss %>% mutate(dss, countInKind = paste(kind,1:n()) )
}
ans <- ds %>% group_by(kind) %>% mapGroups(fSub)
ans
```

Compare this to the nest-map-unnest idiom:
```{r}
ans <- ds %>% group_by(kind) %>% nest() %>% 
  mutate(data = purrr::map2(data, kind, function(dss,kind){
  dss %>% mutate(
    countInKind = paste(first(kind),1:n())
  )
})) %>% unnest()
ans
```



See the [package vignettes](https://github.com/bgctw/dplyrUtil/tree/master/vignettes) (*.md) for  examples.
