---
author: "Thomas Wutzler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Example using equations and figures}
  %\usepackage[UTF-8]{inputenc}
---

Common tasks when using dplyr
=============================

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra = 'style="display:block; margin: auto"'
    #, fig.align = "center"
    , fig.width = 4.3, fig.height = 3.2, dev.args = list(pointsize = 10)
    )
knit_hooks$set(spar = function(before, options, envir) {
    if (before) {
        par( las = 1 )                   #also y axis labels horizontal
        par(mar = c(2.0,3.3,0,0) + 0.3 ) #margins
        par(tck = 0.02 )                 #axe-tick length inside plots             
        par(mgp = c(1.1,0.2,0) )  #positioning of axis title, axis labels, axis
     }
})
```


Introduction
------------

The [tidyverse](https://r4ds.had.co.nz/transform.html) 
with the dplyr package in combination 
with purrr mapping and tidyr data.frames
is agreat way of managing data in R.

Using the package, however, require repeated coding for some common issues.
This package provides routines for some of those issues.

```{r message=FALSE}
require(dplyr, quietly = TRUE)
require(dplyrUtil, quietly = TRUE)
```

split-map-combine
-----------------
One common operation is the [split-map-combine](TODO) pattern 
A tbl, i.e. tabular data such as a data.frame or a tibble,  is
subset, then for each subset a function is called that returns a tbl, and
finally the resulting tbls are row-bind again.

While the tidyverse favours the [group_by-nest-map-unnest](TODO) idiom
of this pattern,
the resulting code is hard to understand for people not familiar
with this idiom. Moreover, the nested data.frames does not 
contain the grouping columns which are sometimes required by
the function working on the subset.

`dplyUtil` provides function `mapGroups` that provides a dplyr like
call interface for the base split function.

The following example applies a function on each kind of food subset
that uses the number of records in the subset and the grouping variable.
```{r}
ds <- tibble(
  food = c("banana","strawberry","pear","bread","corn flakes")
  , kind = factor(c(rep("fruit",3), rep("cereal",2)))
  )
ans <- ds %>% group_by(kind) %>% mapGroups(function(dss){
  dss %>% mutate(
    countInKind = paste(first(kind),1:n())
  )
})
ans
```

If the data.frames to combine hold inconsistent factor levels, by default
a warning is shown and the column is converted to character. 
If instead, one wants to combine all factor levels, there is argument 
`.isFactorReleveled`

```{r}
ans <- ds %>% group_by(kind) %>% mapGroups(function(dss){
  dss %>% mutate(
    countInKind = factor(paste(first(kind),1:n()))
  )
}
, .isFactorReleveled = TRUE
, .noWarningCols = c("countInKind"))
ans
```

mutate acting only on the rows satisfying the condition
-------------------------------------------------------

Default `mutate` acts on all rows. For each modified variable one needs to
use `ifelse`. Function `mutate_cond` helps in those cases.

The following example modifies the Petal.Length column, but only for
a given Species.

```{r}
ans <- iris %>% as_tibble() %>% 
  mutate_cond(
    Species == "setosa"
    , Petal.Length = 1.0
  )
rbind(
ans %>% filter(Species == "setosa") %>% 
  select(Species, Petal.Length) %>% head(3) # 1.0
, ans %>% filter(Species != "setosa") %>% 
  select(Species, Petal.Length) %>% head(3) # orig
) 
```

Joining with Replacement
------------------------

A join is often used to add new columns to a dataset.
If the same information is joined several times, by default the
column is creted with a modified name.
If one instead wants to replace the joined column, function
`left_joinReplace` helps with that.

The following example joines the info column again, 
because it was modified after the first join.
It demonstrates the different results between `left_join` and
`left_joinReplace`.

```{r, output = "hold"}
kindInfo <- tibble(
  kind = factor(c("fruit","cereal"))
  ,info = c("rich in vitamins","rich in carbohydrates")
)
ans1 <- ds %>% left_join(kindInfo, by = "kind")
kindInfo$info[1] <- "juicy"
ans1 %>% left_join(kindInfo, by = "kind") # duplicated info column
ans1 %>% left_joinReplace(kindInfo, by = "kind") # modified info column
```

Dealing with unequal factor levels
----------------------------------

In case of unequal factor levels in row-binding tbls or joins,
factors are converted to character.
Often one wants to combine the factor levels instead.
Function `expandAllInconsistentFactorLevels` helps with that.

In the following example, the levels of factor `f` do not match.
Therefore, `bind_rows` warns and coerces to character.
With expanding the factors before, the combined data.frame factor `f`
is still a factor but with modified levels.

```{r}
df1 <- tibble(f = factor(c("D","D","C")))
df2 <- tibble(f = factor(c("C","C","A"))
                  , desc = c("forC1","forC2","forA1"))
bind_rows(df1,df2) # f is converted to character
expandAllInconsistentFactorLevels(df1,df2) %>% bind_rows()
```

Function `expandFactorLevels` helps to relevel a specific column 
across several tbls.

Both functions proviede the result as a list of
tbls. This type of result can be handled by `bind_rows`, 
but other functions require
single data.frames, i.e. the components of the result (`[[1]]`, `[[2]]`).

For convenience function `left_joinFactors` helps joining 
tbls with different factor levels.
```{r}
left_joinFactors(df1,df2, by = "f", .noWarningCols = c("f"))
```


